---
title: "Analysis & Visualization"
description: "Explore and visualize simulation results with joshpy"
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(reticulate)
library(ggplot2)
library(dplyr)
```

## Introduction

This demo focuses on **exploring and visualizing simulation results** after they've
been collected. The analysis layer in joshpy is **decoupled from orchestration** -
it only needs a `RunRegistry` containing data. Whether the data came from
`SweepManager`, manual workflow, or someone else's experiment doesn't matter.

**Prerequisites**: Run either [Manual Workflow](manual-workflow.qmd) or 
[SweepManager Workflow](sweep-manager.qmd) first to generate `demo_registry.duckdb`.

For orchestration workflows, see:

- [Manual Workflow](manual-workflow.qmd) - Step-by-step control using individual components
- [SweepManager Workflow](sweep-manager.qmd) - Simplified workflow with `SweepManager`

This demo covers:

1. **Data Discovery** - What's in the registry?
2. **Diagnostic Plots** - Quick matplotlib visualizations
3. **Custom Queries** - `DiagnosticQueries` for common patterns
4. **Direct SQL** - Full access to DuckDB for advanced analysis
5. **R/ggplot2** - Publication-quality figures

## Setup: Load Registry

Load the registry saved by a previous orchestration demo:

```{python}
from joshpy.registry import RunRegistry

# Load registry from disk (created by manual-workflow.qmd or sweep-manager.qmd)
REGISTRY_PATH = "demo_registry.duckdb"
registry = RunRegistry(REGISTRY_PATH)
```

## Part 1: Data Discovery

Before plotting, discover what data is available in the registry.

### Full Summary

```{python}
summary = registry.get_data_summary()
print(summary)
```

### Specific Queries

```{python}
# What variables were exported from the simulation?
registry.list_export_variables()
```

```{python}
# What parameters were swept?
registry.list_config_parameters()
```

```{python}
# What entity types are in the data?
registry.list_entity_types()
```

### Variable Columns

```{python}
# List all variable columns in the cell_data table
registry.list_variable_columns()
```

## Part 2: Diagnostic Plots (SimulationDiagnostics)

The `SimulationDiagnostics` class provides quick matplotlib-based visualizations
for simulation sanity checks. These are designed for rapid exploration, not
publication.

```{python}
from joshpy.diagnostics import SimulationDiagnostics

diag = SimulationDiagnostics(registry)
```

### Time Series

Plot how a variable evolves over simulation steps. By default, spatially
aggregates across patches and shows uncertainty bands across replicates.

```{python}
#| label: fig-timeseries-single
#| fig-cap: "Tree height over time for maxGrowth=50, with replicate uncertainty."

# Filter by parameter value
diag.plot_timeseries(
    "averageHeight",
    maxGrowth=50,
    title="Average Tree Height Over Time (maxGrowth=50)",
)
```

#### Aggregation Options

```{python}
#| label: fig-timeseries-aggregates
#| fig-cap: "Different spatial aggregation methods."
#| layout-ncol: 2
#| fig-height: 4

# Sum across patches
diag.plot_timeseries("averageHeight", maxGrowth=50, aggregate="sum", title="Sum")

# Min across patches
diag.plot_timeseries("averageHeight", maxGrowth=50, aggregate="min", title="Min")
```

### Parameter Comparison

Compare a variable across different parameter values - ideal for sweep results.

```{python}
#| label: fig-comparison-lines
#| fig-cap: "Tree height trajectories across all maxGrowth values."

diag.plot_comparison(
    "averageHeight",
    group_by="maxGrowth",
    title="Tree Height by Growth Rate Parameter",
)
```

#### Bar Chart at Specific Step

```{python}
#| label: fig-comparison-bar
#| fig-cap: "Final tree height (step 10) for each maxGrowth value."

diag.plot_comparison(
    "averageHeight",
    group_by="maxGrowth",
    step=10,
    title="Final Tree Height vs Growth Rate",
)
```

### Spatial Snapshot

Scatter plot colored by value at a specific timestep:

```{python}
#| label: fig-spatial
#| fig-cap: "Spatial distribution of tree height at step 10 for maxGrowth=50."

# Get a run hash for maxGrowth=50
run_hash_50 = registry.get_runs_by_parameters(maxGrowth=50)[0]["run_hash"]

diag.plot_spatial(
    "averageHeight",
    step=10,
    run_hash=run_hash_50,
    title="Spatial Distribution (maxGrowth=50, step=10)",
)
```

### Saving Figures

All plot methods return a matplotlib Figure that can be saved:

```{python}
fig = diag.plot_comparison(
    "averageHeight",
    group_by="maxGrowth",
    show=False,  # Don't display inline
)
fig.savefig("/tmp/comparison_plot.png", dpi=150, bbox_inches="tight")
```

## Part 3: Custom Queries (DiagnosticQueries)

For more control than `SimulationDiagnostics`, use `DiagnosticQueries` directly.
These return pandas DataFrames for further processing.

```{python}
from joshpy.cell_data import DiagnosticQueries

queries = DiagnosticQueries(registry)
```

### Parameter Comparison

```{python}
df = queries.get_parameter_comparison(
    variable="averageHeight",
    param_name="maxGrowth",
)
df.head(15)
```

### Replicate Uncertainty

Get statistics across replicates for a specific run:

```{python}
df = queries.get_replicate_uncertainty(
    variable="averageHeight",
    run_hash=run_hash_50,
)
df.head(10)
```

### Spatial Snapshot

Get all spatial data for a given timestep:

```{python}
df = queries.get_spatial_snapshot(
    step=10,
    variable="averageHeight",
    run_hash=run_hash_50,
)
df.head(10)
```

### Get All Variables at Once

Retrieve all exported variables for a specific step:

```{python}
df = queries.get_all_variables_at_step(
    step=10,
    run_hash=run_hash_50,
)
df.head(5)
```

## Part 4: Direct SQL Queries

For full flexibility, query DuckDB directly.

### Simple Aggregations

```{python}
# Mean, min, max across all cells per step
result = registry.query("""
    SELECT 
        step,
        AVG(averageHeight) as mean_height,
        MIN(averageHeight) as min_height,
        MAX(averageHeight) as max_height,
        STDDEV(averageHeight) as std_height,
        COUNT(*) as n_cells
    FROM cell_data
    GROUP BY step
    ORDER BY step
""")
result.df()
```

### Filtering on Variable Values

```{python}
# Find cells where trees grew above a threshold
result = registry.query("""
    SELECT 
        step,
        COUNT(*) as n_cells_above_50,
        AVG(averageHeight) as mean_height_of_those
    FROM cell_data
    WHERE averageHeight > 50
    GROUP BY step
    ORDER BY step
""")
result.df()
```

### Percentile Analysis

```{python}
# Get distribution statistics per step
result = registry.query("""
    SELECT 
        step,
        APPROX_QUANTILE(averageHeight, 0.25) as p25,
        APPROX_QUANTILE(averageHeight, 0.50) as median,
        APPROX_QUANTILE(averageHeight, 0.75) as p75,
        APPROX_QUANTILE(averageHeight, 0.95) as p95
    FROM cell_data
    GROUP BY step
    ORDER BY step
""")
result.df()
```

### Joining with Config Parameters

Config parameters are stored as typed columns in the `config_parameters` table:

```{python}
# Compare results across different parameter values
result = registry.query("""
    SELECT 
        cp.maxGrowth,
        cd.step,
        AVG(cd.averageHeight) as mean_height,
        STDDEV(cd.averageHeight) as std_height
    FROM cell_data cd
    JOIN config_parameters cp ON cd.run_hash = cp.run_hash
    GROUP BY cp.maxGrowth, cd.step
    ORDER BY cp.maxGrowth, cd.step
""")
result.df().head(15)
```

```{python}
# Filter by parameter value (maxGrowth > 50)
result = registry.query("""
    SELECT 
        cd.step,
        AVG(cd.averageHeight) as mean_height
    FROM cell_data cd
    JOIN config_parameters cp ON cd.run_hash = cp.run_hash
    WHERE cp.maxGrowth > 50
    GROUP BY cd.step
    ORDER BY cd.step
""")
result.df()
```

### Export to File

```{python}
# Export query results to CSV
registry.query("""
    SELECT step, AVG(averageHeight) as mean_height
    FROM cell_data
    GROUP BY step
    ORDER BY step
""").df().to_csv("/tmp/height_by_step.csv", index=False)
```

## Part 5: R/ggplot2 Visualization

For publication-quality figures, there are **three routes** to get data from 
joshpy into R:

1. **CSV Export** - Simple, always works
2. **Direct DuckDB** - Custom SQL queries from R, no Python needed
3. **`py$df` via reticulate** - Zero-copy Python/R interop

### Route 1: Export to CSV (Simple, Always Works)

```{python}
# Python exports to CSV
csv_df = queries.get_parameter_comparison(
    variable="averageHeight",
    param_name="maxGrowth",
)
csv_df.to_csv("/tmp/sweep_results.csv", index=False)
```

```{r}
# R reads CSV
df_csv <- read.csv("/tmp/sweep_results.csv")
df_csv$param_value <- as.numeric(df_csv$param_value)
head(df_csv, 5)
```

**Pros**: Always works, simple, portable  
**Cons**: File I/O overhead, temp file management

### Route 2: Direct DuckDB from R (Custom Queries)

R can query DuckDB directly with full SQL flexibility:

```{r}
#| label: r-duckdb-query
library(DBI)
library(duckdb)

con <- dbConnect(duckdb(), "demo_registry.duckdb", read_only = TRUE)

df_duckdb <- dbGetQuery(con, "
    SELECT 
        cp.maxGrowth as param_value,
        cd.step,
        AVG(cd.averageHeight) as mean_value,
        STDDEV(cd.averageHeight) as std_value
    FROM cell_data cd
    JOIN config_parameters cp ON cd.run_hash = cp.run_hash
    GROUP BY cp.maxGrowth, cd.step
    ORDER BY cp.maxGrowth, cd.step
")

dbDisconnect(con, shutdown = TRUE)
head(df_duckdb, 10)
```

**Pros**: Full SQL flexibility, no Python needed for custom queries  
**Cons**: Requires `r-duckdb` package (linux-64 only in pixi)

### Route 3: `py$df` via reticulate (Zero-Copy)

In theory, you can access Python DataFrames directly via `py$df`:

```r
# This should work:
# df <- reticulate::py_to_r(py$sweep_df)
```

We're working on this - for now, Routes 1 and 2 are recommended.

### Plotting with ggplot2

The following examples use `df_duckdb` from Route 2 (direct DuckDB query).
We assign it to `df` for cleaner ggplot code:

```{r}
# Use the DuckDB query result for plotting
df <- df_duckdb
```

### Time Series with Uncertainty Ribbon

```{r}
#| label: fig-r-timeseries
#| fig-cap: "Tree height trajectories with uncertainty ribbons (ggplot2)."
#| fig-width: 10
#| fig-height: 6

ggplot(df, aes(x = step, y = mean_value, 
                     color = factor(param_value), 
                     fill = factor(param_value))) +
  geom_ribbon(aes(ymin = mean_value - std_value, 
                  ymax = mean_value + std_value), 
              alpha = 0.2, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    x = "Simulation Step",
    y = "Average Tree Height (meters)",
    title = "Tree Growth Over Time by maxGrowth Parameter",
    color = "maxGrowth",
    fill = "maxGrowth"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

### Bar Chart with Error Bars

```{r}
#| label: fig-r-barchart
#| fig-cap: "Final tree height vs growth rate parameter with error bars."
#| fig-width: 10
#| fig-height: 6

final_df <- df |>
  dplyr::filter(step == max(step))

ggplot(final_df, aes(x = param_value, y = mean_value)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_errorbar(
    aes(ymin = mean_value - std_value, ymax = mean_value + std_value),
    width = 3,
    linewidth = 0.6
  ) +
  labs(
    x = "maxGrowth (meters/step)",
    y = "Final Average Height (meters)",
    title = "Final Tree Height vs Growth Rate"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

### Faceted Plot

```{r}
#| label: fig-r-faceted
#| fig-cap: "Faceted time series - one panel per maxGrowth value."
#| fig-width: 12
#| fig-height: 8

ggplot(df, aes(x = step, y = mean_value)) +
  geom_ribbon(aes(ymin = mean_value - std_value, 
                  ymax = mean_value + std_value), 
              alpha = 0.3, fill = "steelblue") +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(color = "steelblue", size = 1.5) +
  facet_wrap(~param_value, ncol = 5, labeller = label_both) +
  labs(
    x = "Simulation Step",
    y = "Average Tree Height (meters)",
    title = "Tree Growth Trajectories by maxGrowth Parameter"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "gray90", color = NA)
  )
```

## Summary

This demo covered the joshpy analysis workflow:

| Task | Tool | Notes |
|------|------|-------|
| **Discovery** | `registry.get_data_summary()` | What's in the registry? |
| **Quick plots** | `SimulationDiagnostics` | matplotlib-based, fast iteration |
| **DataFrames** | `DiagnosticQueries` | Returns pandas for further processing |
| **Full SQL** | `registry.query()` | Direct DuckDB access |
| **R (CSV)** | `read.csv()` | Simple, always works |
| **R (DuckDB)** | `dbConnect(duckdb(), ...)` | Custom SQL directly from R |
| **R (reticulate)** | `py$df` | Zero-copy Python/R interop |

**Key Points:**

- Analysis is **decoupled from orchestration** - same workflow for any data source
- Export variables are typed columns in `cell_data`
- Config parameters are typed columns in `config_parameters`
- R can query DuckDB directly
- Three routes for R: CSV export, direct DuckDB, or `py$df` via reticulate

## Cleanup

```{python}
registry.close()
```
