---
title: "Single Run & Iteration"
description: "Run a simulation, tweak parameters, compare results: the daily research workflow"
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(reticulate)
library(ggplot2)
library(dplyr)
```

## Overview

This tutorial walks through the most common joshpy workflow:

1. **Run** a simulation with a config file
2. **Inspect** the results
3. **Tweak** a parameter and run again
4. **Compare** the two runs side by side

This is the "tinkering" workflow researchers use day-to-day before committing to
a systematic parameter sweep.

## Prerequisites

```bash
pip install -e '.[all]'
```

## The Example Model

We use the tutorial tree growth model, which tracks tree height and age over time.
The model reads two config parameters:

- `initialTreeCount` -- how many trees per patch at initialization
- `maxGrowth` -- maximum height growth per timestep (meters)

```{python}
from pathlib import Path

SOURCE_PATH = Path("../../examples/tutorial_sweep.josh")
print(SOURCE_PATH.read_text())
```

## Step 1: First Run (Baseline)

Create a `JobConfig` pointing at the model and a raw `.jshc` config file.
When you use `config_path` (a raw `.jshc`, no Jinja templating), joshpy
**auto-parses** all parameter values and stores them in the registry for
querying.

```{python}
BASELINE_CONFIG = Path("../../examples/configs/baseline.jshc")
print(BASELINE_CONFIG.read_text())
```

```{python}
from joshpy.jobs import JobConfig
from joshpy.sweep import SweepManager
from joshpy.cli import JoshCLI
from joshpy.jar import JarMode
from joshpy.catalog import ProjectCatalog

REGISTRY_PATH = "iteration_demo.duckdb"
catalog = ProjectCatalog("catalog_demo.duckdb")

config = JobConfig(
    source_path=SOURCE_PATH,
    config_path=BASELINE_CONFIG,
    simulation="Main",
    replicates=3,
    label="baseline",
)

cli = JoshCLI(josh_jar=JarMode.DEV)

manager = (
    SweepManager.builder(config)
    .with_registry(REGISTRY_PATH, experiment_name="baseline")
    .with_cli(cli)
    .with_label("baseline")
    .with_catalog(catalog, experiment_name="iteration_demo")
    .build()
)

try:
    results = manager.run()
    print(f"Completed: {results.succeeded} succeeded, {results.failed} failed")

    if results.failed > 0:
        for job, result in results:
            if not result.success:
                print(f"Error: {result.stderr[:300]}")
        raise RuntimeError("Run failed")

    manager.load_results()
finally:
    manager.cleanup()
    manager.close()
```

The `label` is registry metadata — a human-readable name stored alongside the
run's hash for querying and comparison. It is not a simulation parameter and does
not affect the run itself. Setting it on `JobConfig` or via `.with_label()` also
makes it available as a `{label}` template variable in Josh export paths.

### Quick Look

```{python}
from joshpy.registry import RunRegistry
from joshpy.diagnostics import SimulationDiagnostics

registry = RunRegistry(REGISTRY_PATH)

summary = registry.get_data_summary()
print(summary)
```

```{python}
diag = SimulationDiagnostics(registry)
diag.plot_timeseries("averageHeight", title="Baseline - Average Height Over Time")
```

```{python}
registry.close()
```

## Step 2: Tweak and Re-Run

What happens if we double `maxGrowth` from 50 to 100? Edit `baseline.jshc`
(or use a second config file) and re-run with the **same label**. The
`on_collision="supersede"` option retires the old run — it releases the label
and is marked `status="superseded"` pointing at the new run — so the bare
`"baseline"` label always points to your latest attempt:

```{python}
HIGH_GROWTH_CONFIG = Path("../../examples/configs/high_growth.jshc")
print(HIGH_GROWTH_CONFIG.read_text())
```

```{python}
config2 = JobConfig(
    source_path=SOURCE_PATH,
    config_path=HIGH_GROWTH_CONFIG,
    simulation="Main",
    replicates=3,
    label="high_growth",
)

manager = (
    SweepManager.builder(config2)
    .with_registry(REGISTRY_PATH, experiment_name="high_growth")
    .with_cli(cli)
    .with_label("baseline", on_collision="supersede", reason="doubled maxGrowth")
    .with_catalog(catalog, experiment_name="iteration_demo")
    .build()
)

try:
    results = manager.run()
    print(f"Completed: {results.succeeded} succeeded, {results.failed} failed")

    if results.failed > 0:
        for job, result in results:
            if not result.success:
                print(f"Error: {result.stderr[:300]}")
        raise RuntimeError("Run failed")

    manager.load_results()
finally:
    manager.cleanup()
    manager.close()
```

Both runs still live in the same DuckDB file, but the old baseline is now
**superseded**: it released the `"baseline"` label (which points to the new run)
and carries `status="superseded"` with a `superseded_by` link back to the new
run. Its results stay queryable for provenance, but "current" reads skip it.

::: {.callout-tip title="Alternatives to on_collision"}
If you prefer explicit control, you can use a separate `.jshc` file with its
own label (e.g., `.with_label("high_growth")`), or call
`registry.label_run(run_hash, "name")` after the fact. Use `force=True` instead
of `on_collision="supersede"` when you simply mislabeled a run and want to move
the label without recording a supersession.
:::

## Step 3: Compare

### See What's There

```{python}
registry = RunRegistry(REGISTRY_PATH)

# Which labels exist?
registry.list_labels()
```

```{python}
# What parameters were auto-parsed from the .jshc files?
registry.list_config_parameters()
```

::: {.callout-note title="Labels vs Parameters"}
Labels are registry metadata — human-readable names you assign for organizing
and querying runs (e.g., `group_by="label"`). Parameters like `maxGrowth` are
the actual simulation inputs extracted from the stored config. A label does not
affect the simulation; it identifies the run for downstream comparison and
analysis.
:::

### Trace the Supersession Lineage

`list_labels()` shows only the current `"baseline"`, because the old run
released its label. To see the full iteration history — current run first, then
each run it superseded — walk the lineage with `run_history()`:

```{python}
for detail in registry.run_history("baseline"):
    status = registry.get_run_status(detail.run_hash)
    print(f"{detail.run_hash}  status={status.effective_status}  "
          f"maxGrowth={detail.parameters.get('maxGrowth')}")
```

### Compare by Parameter Value

Currency retires the *label*, not the *data*: both runs' results are still in
the registry, so you can compare them by the parameter that actually changed.
Since `.jshc` parameters were auto-parsed, group by that parameter directly:

```{python}
diag = SimulationDiagnostics(registry)
```

```{python}
#| label: fig-param-comparison
#| fig-cap: "Average tree height grouped by the maxGrowth config value."

diag.plot_comparison(
    "averageHeight",
    group_by="maxGrowth",
    title="Average Height by maxGrowth Parameter",
)
```

### Inspect Stored Configs

The registry stores the full config content for every run, so you can always
review or diff what was run -- even if the original files have changed. See
[Inspecting Runs from the Registry](inspecting-runs.qmd) for the full
view/diff workflow, including the `python -m joshpy.inspect` CLI.

## Analysis in R

The registry is a standard DuckDB file. R can connect directly and query labels:

```{r}
#| label: fig-r-comparison
#| fig-cap: "Same comparison in R/ggplot2, querying the DuckDB registry directly."
#| fig-width: 10
#| fig-height: 5

library(DBI)
library(duckdb)

con <- dbConnect(duckdb::duckdb(), "iteration_demo.duckdb", read_only = TRUE)

df <- dbGetQuery(con, "
    SELECT jc.label, cd.step,
           AVG(cd.averageHeight) AS mean_height,
           STDDEV(cd.averageHeight) AS sd_height
    FROM cell_data cd
    JOIN job_configs jc ON cd.run_hash = jc.run_hash
    WHERE jc.label IS NOT NULL
    GROUP BY jc.label, cd.step
    ORDER BY jc.label, cd.step
")

ggplot(df, aes(x = step, y = mean_height, color = label, fill = label)) +
  geom_ribbon(aes(ymin = pmax(mean_height - sd_height, 0),
                  ymax = mean_height + sd_height),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(x = "Simulation Step", y = "Average Height (meters)",
       title = "Tree Height: Baseline Iterations",
       color = "Run", fill = "Run") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

dbDisconnect(con, shutdown = TRUE)
```

## Key Concepts

### Labels

Labels are human-readable names for individual runs. Set them upfront with
`.with_label()` on the builder. They're unique within a registry.

- **Single-job configs**: `.with_label("name")` works at build time
- **Multi-job sweeps**: Label individual runs after the fact with
  `registry.label_run(run_hash, "name")`
- **Re-running with the same label** (shown in Step 2): `on_collision="supersede"`
  retires the old holder — it releases the label and is marked
  `status="superseded"` pointing at the new run — so `resolve_label("baseline")`
  always returns the current run. Walk the lineage with
  `registry.run_history("baseline")` (current first, then each superseded run).
- **Force-reassigning a label**: `force=True` silently drops the old label
  without recording supersession — for the "I mislabeled it, just fix it" case.

### Run status (currency)

Every run has a lifecycle `status` — `active` (the default; also what an unmarked
run reads as), `superseded`, or `bad`. "Current" is exactly `status = "active"`.
Mark runs explicitly:

- `registry.mark_run(run_hash, "bad", reason="...")` — exclude a run you no
  longer trust.
- `registry.supersede(new_hash, replaces="baseline", reason="...")` — record that
  `new_hash` replaced an existing run (releasing its label).
- `registry.get_run_status(label_or_hash)` — inspect a run's status,
  `superseded_by` link, and reason.

The `cell_data_current` view (`SELECT ... FROM cell_data_current`) restricts
local SQL to rows from active runs only.

### Auto-Parsed Config Parameters

When you use `config_path` (raw `.jshc`, no Jinja template), joshpy auto-parses
all `name = value unit` lines and stores them as typed parameter columns. This
enables `group_by="maxGrowth"` without defining a sweep.

### Same Registry, Multiple Runs

All runs that share a registry are queryable together. Keep pointing at the same
`.duckdb` file and results accumulate.

### ProjectCatalog (for multi-session tracking)

The `.with_catalog()` lines in Steps 1 and 2 registered each run in a
`ProjectCatalog` -- a persistent DuckDB database that indexes your experiments
across sessions. The catalog tracks which models, configs, and data were used,
where results live, and whether an experiment was already run. It does NOT store
simulation data -- that stays in per-experiment registry files.

When you come back to a project after days or weeks, the catalog tells you
what's been done:

```{python}
experiments = catalog.list_experiments()
print(f"Total experiments: {len(experiments)}\n")
for exp in experiments:
    print(f"  {exp.name}: status={exp.status}, registry={exp.registry_path}")
```

Before running a new experiment, check if it's already been done:

```{python}
existing = catalog.find_experiment(config)
if existing:
    print(f"Already run: '{existing.name}' at {existing.registry_path}")
else:
    print("Not yet run")
```

| | ProjectCatalog | RunRegistry |
|---|---|---|
| **Stores** | Hashes and pointers | Simulation data |
| **Scope** | One per project | One per experiment |
| **Persistence** | Across all sessions | Across sessions for one experiment |
| **Primary use** | "What have I run?" | "What were the results?" |

::: {.callout-note collapse="true" title="Advanced: Cross-registry queries"}
If different researchers maintain separate registries (or you used separate
registries for each run), `catalog.open_registries()` can combine them via
DuckDB ATTACH:

```python
experiments = catalog.list_experiments(status="completed")

with catalog.open_registries(experiments) as conn:
    parts = []
    for i, exp in enumerate(experiments):
        parts.append(
            f"SELECT '{exp.name}' as experiment, step, "
            f"AVG(averageHeight) as mean_height "
            f"FROM exp_{i}.cell_data GROUP BY step"
        )
    sql = " UNION ALL ".join(parts) + " ORDER BY experiment, step"
    df = conn.execute(sql).df()
```
:::

### GridSpec (for projects with external data)

GridSpec is not required for basic iteration -- you can pass `file_mappings` as
a plain dict or omit it entirely if your model has no external data. But when
your project has preprocessed `.jshd` files across multiple scenarios, GridSpec
keeps the inventory organized and scales to sweeps later.

If your model uses external data files, a `GridSpec` YAML file defines grid
geometry and inventories the data files:

```yaml
# data/grids/dev_fine/grid.yaml
name: dev_fine
grid:
  size_m: 30
  low: [33.902, -116.0465]
  high: [33.908, -116.0395]

time:
  type: count
  start: 2015
  unit: year
  count: 86
  increment: 1

variants:
  scenario:
    values: [ssp245, ssp370, ssp585]
    default: ssp245

files:
  cover:
    path: cover.jshd
    units: percent
  futureTempJan:
    template_path: monthly/tas_{scenario}_jan.jshd
    units: K
```

```python
from joshpy.grid import GridSpec

grid = GridSpec.from_yaml("data/grids/dev_fine/grid.yaml")

# Single run with a specific scenario
config = JobConfig(
    source_path=Path("model.josh"),
    config_path=Path("config.jshc"),
    file_mappings=grid.file_mappings_for(scenario="ssp370"),
)
```

See [Project Organization](project-organization.qmd) for the full GridSpec guide
including variant data files and sweep generation.

## Next Steps

- [SweepManager Workflow](sweep-manager.qmd) -- Systematic parameter sweeps
  when you know what to sweep
- [Multi-Session Sweeps](multi-session-sweeps.qmd) -- Catalog as bookkeeper for parameter sweeps
- [Analysis & Visualization](analysis.qmd) -- Deeper dive into queries, spatial
  plots, and R integration
- [Project Organization](project-organization.qmd) -- Multi-grid, multi-model
  project structure with GridSpec

## Cleanup

```{python}
import os

registry.close()
catalog.close()

for f in ["iteration_demo.duckdb", "iteration_demo.duckdb.wal",
          "catalog_demo.duckdb"]:
    if os.path.exists(f):
        os.remove(f)
```
