---
title: "Single Run & Iteration"
description: "Run a simulation, tweak parameters, compare results: the daily research workflow"
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(reticulate)
library(ggplot2)
library(dplyr)
```

## Overview

This tutorial walks through the most common joshpy workflow:

1. **Run** a simulation with a config file
2. **Inspect** the results
3. **Tweak** a parameter and run again
4. **Compare** the two runs side by side

This is the "tinkering" workflow researchers use day-to-day before committing to
a systematic parameter sweep.

## Prerequisites

```bash
pip install -e '.[all]'
```

## The Example Model

We use the tutorial tree growth model, which tracks tree height and age over time.
The model reads two config parameters:

- `initialTreeCount` -- how many trees per patch at initialization
- `maxGrowth` -- maximum height growth per timestep (meters)

```{python}
from pathlib import Path

SOURCE_PATH = Path("../../examples/tutorial_sweep.josh")
print(SOURCE_PATH.read_text())
```

## Step 1: First Run (Baseline)

Create a `JobConfig` pointing at the model and a raw `.jshc` config file.
When you use `config_path` (a raw `.jshc`, no Jinja templating), joshpy
**auto-parses** all parameter values and stores them in the registry for
querying.

```{python}
BASELINE_CONFIG = Path("../../examples/configs/baseline.jshc")
print(BASELINE_CONFIG.read_text())
```

```{python}
from joshpy.jobs import JobConfig
from joshpy.sweep import SweepManager
from joshpy.cli import JoshCLI
from joshpy.jar import JarMode
from joshpy.catalog import ProjectCatalog

REGISTRY_PATH = "iteration_demo.duckdb"
catalog = ProjectCatalog("catalog_demo.duckdb")

config = JobConfig(
    source_path=SOURCE_PATH,
    config_path=BASELINE_CONFIG,
    simulation="Main",
    replicates=3,
)

cli = JoshCLI(josh_jar=JarMode.DEV)

manager = (
    SweepManager.builder(config)
    .with_registry(REGISTRY_PATH, experiment_name="baseline")
    .with_cli(cli)
    .with_label("baseline")
    .with_catalog(catalog, experiment_name="iteration_demo")
    .build()
)

try:
    results = manager.run()
    print(f"Completed: {results.succeeded} succeeded, {results.failed} failed")

    if results.failed > 0:
        for job, result in results:
            if not result.success:
                print(f"Error: {result.stderr[:300]}")
        raise RuntimeError("Run failed")

    manager.load_results()
finally:
    manager.cleanup()
    manager.close()
```

The label `"baseline"` is set **upfront** at build time via `.with_label()`,
not after the run. This avoids forgetting to label.

### Quick Look

```{python}
from joshpy.registry import RunRegistry
from joshpy.diagnostics import SimulationDiagnostics

registry = RunRegistry(REGISTRY_PATH)

summary = registry.get_data_summary()
print(summary)
```

```{python}
diag = SimulationDiagnostics(registry)
diag.plot_timeseries("averageHeight", title="Baseline - Average Height Over Time")
```

```{python}
registry.close()
```

## Step 2: Tweak and Re-Run

What happens if we double `maxGrowth` from 50 to 100? We have a second config
file with just that change:

```{python}
HIGH_GROWTH_CONFIG = Path("../../examples/configs/high_growth.jshc")
print(HIGH_GROWTH_CONFIG.read_text())
```

Run again, **pointing at the same registry**:

```{python}
config2 = JobConfig(
    source_path=SOURCE_PATH,
    config_path=HIGH_GROWTH_CONFIG,
    simulation="Main",
    replicates=3,
)

manager = (
    SweepManager.builder(config2)
    .with_registry(REGISTRY_PATH, experiment_name="high_growth")
    .with_cli(cli)
    .with_label("high_growth")
    .with_catalog(catalog, experiment_name="iteration_demo")
    .build()
)

try:
    results = manager.run()
    print(f"Completed: {results.succeeded} succeeded, {results.failed} failed")

    if results.failed > 0:
        for job, result in results:
            if not result.success:
                print(f"Error: {result.stderr[:300]}")
        raise RuntimeError("Run failed")

    manager.load_results()
finally:
    manager.cleanup()
    manager.close()
```

Both runs now live in the same DuckDB file. Each has a unique label and a
content-based hash.

::: {.callout-tip title="Single-File Workflow"}
You don't need separate `.jshc` files for each run. You can edit
`baseline.jshc` directly -- change `maxGrowth` from 50 to 100 -- and re-run
with `.with_label("high_growth")`. The registry stores the full config content
at registration time, so both versions are preserved even though only one file
exists on disk. The run hash changes automatically because the content changed.
:::

## Step 3: Compare

### See What's There

```{python}
registry = RunRegistry(REGISTRY_PATH)

# Which labels exist?
registry.list_labels()
```

```{python}
# What parameters were auto-parsed from the .jshc files?
registry.list_config_parameters()
```

::: {.callout-note title="Labels vs Parameters"}
`group_by="label"` groups by user-assigned names -- organizational metadata you
control. `group_by="maxGrowth"` groups by the actual parameter value extracted
from the stored config content. Labels carry whatever meaning you give them;
parameter grouping uses the run's own data.
:::

### Compare by Label

One line per labeled run:

```{python}
#| label: fig-label-comparison
#| fig-cap: "Average tree height: baseline (maxGrowth=50) vs high growth (maxGrowth=100)."

diag = SimulationDiagnostics(registry)

diag.plot_comparison(
    "averageHeight",
    group_by="label",
    title="Baseline vs High Growth",
)
```

### Compare by Parameter Value

Since `.jshc` parameters were auto-parsed, you can also group by the parameter
that actually changed:

```{python}
#| label: fig-param-comparison
#| fig-cap: "Average tree height grouped by the maxGrowth config value."

diag.plot_comparison(
    "averageHeight",
    group_by="maxGrowth",
    title="Average Height by maxGrowth Parameter",
)
```

### Export Configs for IDE Diffing

To see exactly what changed between two runs, export the stored configs:

```{python}
import tempfile
tmpdir = tempfile.mkdtemp()

path1 = registry.export_config("baseline", tmpdir)
path2 = registry.export_config("high_growth", tmpdir)

print(f"Exported: {path1.name}, {path2.name}")
print(f"\n--- {path1.name} ---")
print(path1.read_text())
print(f"--- {path2.name} ---")
print(path2.read_text())
```

In practice you'd diff these in your IDE:

```bash
code --diff tmp/baseline.jshc tmp/high_growth.jshc
```

The configs are stored in the registry at registration time, so this works even
if you later modify or delete the original `.jshc` files.

::: {.callout-tip title="Command-Line Config Tools"}
You can view or diff configs directly from the terminal:

```bash
# View a single config
python -m joshpy.config iteration_demo.duckdb --view baseline

# Diff two configs in VS Code
python -m joshpy.config iteration_demo.duckdb --diff baseline high_growth
```

See `python -m joshpy.config --help` for options (IDE selection, export-only mode).
:::

## Analysis in R

The registry is a standard DuckDB file. R can connect directly and query labels:

```{r}
#| label: fig-r-comparison
#| fig-cap: "Same comparison in R/ggplot2, querying the DuckDB registry directly."
#| fig-width: 10
#| fig-height: 5

library(DBI)
library(duckdb)

con <- dbConnect(duckdb::duckdb(), "iteration_demo.duckdb", read_only = TRUE)

df <- dbGetQuery(con, "
    SELECT jc.label, cd.step,
           AVG(cd.averageHeight) AS mean_height,
           STDDEV(cd.averageHeight) AS sd_height
    FROM cell_data cd
    JOIN job_configs jc ON cd.run_hash = jc.run_hash
    WHERE jc.label IS NOT NULL
    GROUP BY jc.label, cd.step
    ORDER BY jc.label, cd.step
")

ggplot(df, aes(x = step, y = mean_height, color = label, fill = label)) +
  geom_ribbon(aes(ymin = pmax(mean_height - sd_height, 0),
                  ymax = mean_height + sd_height),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(x = "Simulation Step", y = "Average Height (meters)",
       title = "Tree Height: Baseline vs High Growth",
       color = "Run", fill = "Run") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

dbDisconnect(con, shutdown = TRUE)
```

## Key Concepts

### Labels

Labels are human-readable names for individual runs. Set them upfront with
`.with_label()` on the builder. They're unique within a registry.

- **Single-job configs**: `.with_label("name")` works at build time
- **Multi-job sweeps**: Label individual runs after the fact with
  `registry.label_run(run_hash, "name")`

### Auto-Parsed Config Parameters

When you use `config_path` (raw `.jshc`, no Jinja template), joshpy auto-parses
all `name = value unit` lines and stores them as typed parameter columns. This
enables `group_by="maxGrowth"` without defining a sweep.

### Same Registry, Multiple Runs

All runs that share a registry are queryable together. Keep pointing at the same
`.duckdb` file and results accumulate.

### ProjectCatalog (for multi-session tracking)

The `.with_catalog()` lines in Steps 1 and 2 registered each run in a
`ProjectCatalog` -- a persistent DuckDB database that indexes your experiments
across sessions. The catalog tracks which models, configs, and data were used,
where results live, and whether an experiment was already run. It does NOT store
simulation data -- that stays in per-experiment registry files.

When you come back to a project after days or weeks, the catalog tells you
what's been done:

```{python}
experiments = catalog.list_experiments()
print(f"Total experiments: {len(experiments)}\n")
for exp in experiments:
    print(f"  {exp.name}: status={exp.status}, registry={exp.registry_path}")
```

Before running a new experiment, check if it's already been done:

```{python}
existing = catalog.find_experiment(config)
if existing:
    print(f"Already run: '{existing.name}' at {existing.registry_path}")
else:
    print("Not yet run")
```

| | ProjectCatalog | RunRegistry |
|---|---|---|
| **Stores** | Hashes and pointers | Simulation data |
| **Scope** | One per project | One per experiment |
| **Persistence** | Across all sessions | Across sessions for one experiment |
| **Primary use** | "What have I run?" | "What were the results?" |

::: {.callout-note collapse="true" title="Advanced: Cross-registry queries"}
If different researchers maintain separate registries (or you used separate
registries for each run), `catalog.open_registries()` can combine them via
DuckDB ATTACH:

```python
experiments = catalog.list_experiments(status="completed")

with catalog.open_registries(experiments) as conn:
    parts = []
    for i, exp in enumerate(experiments):
        parts.append(
            f"SELECT '{exp.name}' as experiment, step, "
            f"AVG(averageHeight) as mean_height "
            f"FROM exp_{i}.cell_data GROUP BY step"
        )
    sql = " UNION ALL ".join(parts) + " ORDER BY experiment, step"
    df = conn.execute(sql).df()
```
:::

### GridSpec (for projects with external data)

GridSpec is not required for basic iteration -- you can pass `file_mappings` as
a plain dict or omit it entirely if your model has no external data. But when
your project has preprocessed `.jshd` files across multiple scenarios, GridSpec
keeps the inventory organized and scales to sweeps later.

If your model uses external data files, a `GridSpec` YAML file defines grid
geometry and inventories the data files:

```yaml
# data/grids/dev_fine/grid.yaml
name: dev_fine
grid:
  size_m: 30
  low: [33.902, -116.0465]
  high: [33.908, -116.0395]
  steps: 86

variants:
  scenario:
    values: [ssp245, ssp370, ssp585]
    default: ssp245

files:
  cover:
    path: cover.jshd
    units: percent
  futureTempJan:
    template_path: monthly/tas_{scenario}_jan.jshd
    units: K
```

```python
from joshpy.grid import GridSpec

grid = GridSpec.from_yaml("data/grids/dev_fine/grid.yaml")

# Single run with a specific scenario
config = JobConfig(
    source_path=Path("model.josh"),
    config_path=Path("config.jshc"),
    file_mappings=grid.file_mappings_for(scenario="ssp370"),
)
```

See [Project Organization](project-organization.qmd) for the full GridSpec guide
including variant data files and sweep generation.

## Next Steps

- [SweepManager Workflow](sweep-manager.qmd) -- Systematic parameter sweeps
  when you know what to sweep
- [Multi-Session Sweeps](multi-session-sweeps.qmd) -- Catalog as bookkeeper for parameter sweeps
- [Analysis & Visualization](analysis.qmd) -- Deeper dive into queries, spatial
  plots, and R integration
- [Project Organization](project-organization.qmd) -- Multi-grid, multi-model
  project structure with GridSpec

## Cleanup

```{python}
import os

registry.close()
catalog.close()

for f in ["iteration_demo.duckdb", "iteration_demo.duckdb.wal",
          "catalog_demo.duckdb"]:
    if os.path.exists(f):
        os.remove(f)
```
