---
title: "Single Run & Iteration"
description: "Run a simulation, tweak parameters, compare results: the daily research workflow"
execute:
  warning: false
---

```{r}
#| label: setup
#| include: false
library(reticulate)
library(ggplot2)
library(dplyr)
```

## Overview

This tutorial walks through the most common joshpy workflow:

1. **Run** a simulation with a config file
2. **Inspect** the results
3. **Tweak** a parameter and run again
4. **Compare** the two runs side by side

This is the "tinkering" workflow researchers use day-to-day before committing to
a systematic parameter sweep.

## Prerequisites

```bash
pip install -e '.[all]'
```

## The Example Model

We use the tutorial tree growth model, which tracks tree height and age over time.
The model reads two config parameters:

- `initialTreeCount` -- how many trees per patch at initialization
- `maxGrowth` -- maximum height growth per timestep (meters)

```{python}
from pathlib import Path

SOURCE_PATH = Path("../../examples/tutorial_sweep.josh")
print(SOURCE_PATH.read_text())
```

## Step 1: First Run (Baseline)

Create a `JobConfig` pointing at the model and a raw `.jshc` config file.
When you use `config_path` (a raw `.jshc`, no Jinja templating), joshpy
**auto-parses** all parameter values and stores them in the registry for
querying.

```{python}
BASELINE_CONFIG = Path("../../examples/configs/baseline.jshc")
print(BASELINE_CONFIG.read_text())
```

```{python}
from joshpy.jobs import JobConfig
from joshpy.sweep import SweepManager
from joshpy.cli import JoshCLI
from joshpy.jar import JarMode

REGISTRY_PATH = "iteration_demo.duckdb"

config = JobConfig(
    source_path=SOURCE_PATH,
    config_path=BASELINE_CONFIG,
    simulation="Main",
    replicates=3,
)

cli = JoshCLI(josh_jar=JarMode.DEV)

manager = (
    SweepManager.builder(config)
    .with_registry(REGISTRY_PATH, experiment_name="baseline")
    .with_cli(cli)
    .with_label("baseline")
    .build()
)

try:
    results = manager.run()
    print(f"Completed: {results.succeeded} succeeded, {results.failed} failed")

    if results.failed > 0:
        for job, result in results:
            if not result.success:
                print(f"Error: {result.stderr[:300]}")
        raise RuntimeError("Run failed")

    manager.load_results()
finally:
    manager.cleanup()
    manager.close()
```

The label `"baseline"` is set **upfront** at build time via `.with_label()`,
not after the run. This avoids forgetting to label.

### Quick Look

```{python}
from joshpy.registry import RunRegistry
from joshpy.diagnostics import SimulationDiagnostics

registry = RunRegistry(REGISTRY_PATH)

summary = registry.get_data_summary()
print(summary)
```

```{python}
diag = SimulationDiagnostics(registry)
diag.plot_timeseries("averageHeight", title="Baseline - Average Height Over Time")
```

```{python}
registry.close()
```

## Step 2: Tweak and Re-Run

What happens if we double `maxGrowth` from 50 to 100? We have a second config
file with just that change:

```{python}
HIGH_GROWTH_CONFIG = Path("../../examples/configs/high_growth.jshc")
print(HIGH_GROWTH_CONFIG.read_text())
```

Run again, **pointing at the same registry**:

```{python}
config2 = JobConfig(
    source_path=SOURCE_PATH,
    config_path=HIGH_GROWTH_CONFIG,
    simulation="Main",
    replicates=3,
)

manager = (
    SweepManager.builder(config2)
    .with_registry(REGISTRY_PATH, experiment_name="high_growth")
    .with_cli(cli)
    .with_label("high_growth")
    .build()
)

try:
    results = manager.run()
    print(f"Completed: {results.succeeded} succeeded, {results.failed} failed")

    if results.failed > 0:
        for job, result in results:
            if not result.success:
                print(f"Error: {result.stderr[:300]}")
        raise RuntimeError("Run failed")

    manager.load_results()
finally:
    manager.cleanup()
    manager.close()
```

Both runs now live in the same DuckDB file. Each has a unique label and a
content-based hash.

## Step 3: Compare

### See What's There

```{python}
registry = RunRegistry(REGISTRY_PATH)

# Which labels exist?
registry.list_labels()
```

```{python}
# What parameters were auto-parsed from the .jshc files?
registry.list_config_parameters()
```

### Compare by Label

One line per labeled run:

```{python}
#| label: fig-label-comparison
#| fig-cap: "Average tree height: baseline (maxGrowth=50) vs high growth (maxGrowth=100)."

diag = SimulationDiagnostics(registry)

diag.plot_comparison(
    "averageHeight",
    group_by="label",
    title="Baseline vs High Growth",
)
```

### Compare by Parameter Value

Since `.jshc` parameters were auto-parsed, you can also group by the parameter
that actually changed:

```{python}
#| label: fig-param-comparison
#| fig-cap: "Average tree height grouped by the maxGrowth config value."

diag.plot_comparison(
    "averageHeight",
    group_by="maxGrowth",
    title="Average Height by maxGrowth Parameter",
)
```

### Export Configs for IDE Diffing

To see exactly what changed between two runs, export the stored configs:

```{python}
import tempfile
tmpdir = tempfile.mkdtemp()

path1 = registry.export_config("baseline", tmpdir)
path2 = registry.export_config("high_growth", tmpdir)

print(f"Exported: {path1.name}, {path2.name}")
print(f"\n--- {path1.name} ---")
print(path1.read_text())
print(f"--- {path2.name} ---")
print(path2.read_text())
```

In practice you'd diff these in your IDE:

```bash
code --diff tmp/baseline.jshc tmp/high_growth.jshc
```

The configs are stored in the registry at registration time, so this works even
if you later modify or delete the original `.jshc` files.

## Analysis in R

The registry is a standard DuckDB file. R can connect directly and query labels:

```{r}
#| label: fig-r-comparison
#| fig-cap: "Same comparison in R/ggplot2, querying the DuckDB registry directly."
#| fig-width: 10
#| fig-height: 5

library(DBI)
library(duckdb)

con <- dbConnect(duckdb::duckdb(), "iteration_demo.duckdb", read_only = TRUE)

df <- dbGetQuery(con, "
    SELECT jc.label, cd.step,
           AVG(cd.averageHeight) AS mean_height,
           STDDEV(cd.averageHeight) AS sd_height
    FROM cell_data cd
    JOIN job_configs jc ON cd.run_hash = jc.run_hash
    WHERE jc.label IS NOT NULL
    GROUP BY jc.label, cd.step
    ORDER BY jc.label, cd.step
")

ggplot(df, aes(x = step, y = mean_height, color = label, fill = label)) +
  geom_ribbon(aes(ymin = pmax(mean_height - sd_height, 0),
                  ymax = mean_height + sd_height),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(x = "Simulation Step", y = "Average Height (meters)",
       title = "Tree Height: Baseline vs High Growth",
       color = "Run", fill = "Run") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())

dbDisconnect(con, shutdown = TRUE)
```

## Key Concepts

### Labels

Labels are human-readable names for individual runs. Set them upfront with
`.with_label()` on the builder. They're unique within a registry.

- **Single-job configs**: `.with_label("name")` works at build time
- **Multi-job sweeps**: Label individual runs after the fact with
  `registry.label_run(run_hash, "name")`

### Auto-Parsed Config Parameters

When you use `config_path` (raw `.jshc`, no Jinja template), joshpy auto-parses
all `name = value unit` lines and stores them as typed parameter columns. This
enables `group_by="maxGrowth"` without defining a sweep.

### Same Registry, Multiple Runs

All runs that share a registry are queryable together. Keep pointing at the same
`.duckdb` file and results accumulate.

### GridSpec (for projects with external data)

If your model uses external data files (`.jshd`), a `GridSpec` YAML file
defines grid geometry and inventories the data files:

```yaml
# data/grids/dev_fine/grid.yaml
name: dev_fine
grid:
  size_m: 30
  low: [33.902, -116.0465]
  high: [33.908, -116.0395]
  steps: 86

files:
  cover:
    path: cover.jshd
    units: percent
```

```python
from joshpy.grid import GridSpec

grid = GridSpec.from_yaml("data/grids/dev_fine/grid.yaml")
config = JobConfig(
    source_path=Path("model.josh"),
    config_path=Path("config.jshc"),
    file_mappings=grid.file_mappings,
)
```

See [Project Organization](project-organization.qmd) for the full GridSpec guide.

## Next Steps

- [SweepManager Workflow](sweep-manager.qmd) -- Systematic parameter sweeps
  when you know what to sweep
- [Analysis & Visualization](analysis.qmd) -- Deeper dive into queries, spatial
  plots, and R integration
- [Project Organization](project-organization.qmd) -- Multi-grid, multi-model
  project structure with GridSpec and ProjectCatalog

## Cleanup

```{python}
import os

registry.close()

for f in ["iteration_demo.duckdb", "iteration_demo.duckdb.wal"]:
    if os.path.exists(f):
        os.remove(f)
```
